---
title: "bearded_CKMR_data_munging"
author: "Paul Conn"
date: "5/2/2022"
output:
  pdf_document: default
  html_document: default
bibliography: master_bib.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document is intended to describe the processes and rationale in transforming bearded seal data provided by ADF&G into a format suitable for analyzing with close-kin mark-recapture (CKMR) models.  These data are almost there, but it will be necessary to (1) remove a few records and reformat data, and (2) formulate posterior distributions for the age of each sample.  We'll describe each of these processes in turn after describing data inputs.

## Source data

The data sets provided by ADF&G include "Samples.csv", "MHS_matches.csv", "PHS_matches.csv", and "POP_matches.csv".  The "Samples.csv" file includes a record for each individual seal, as well as sex, age data, date of kill, village of harvest, and some information from the genetic analysis  The other files provide row indices for maternal half-sibling, paternal half-sibling, and parent-offspring pairs, respectively, as well as a "PLOD" log odds score.  The PLOD score was used in kin finding and may be used to eliminate certain half-sibling pairs from the match list, in the event we try to use a higher PLOD threshold to discriminate matches from non-matches.

## Eliminating individuals

There are several individuals included in "Samples.csv" that we'll likely need to remove from analysis.  However, this means that the row numbers of all individuals below the removed animals in the Samples dataset will change, and we'll want to be able to adjust the various row indices in the various matches files to match updated Sample datasets.  Let's provide a framework for this by attaching individual identifiers to each of the "matches" datasets.  We'll then remove some duplicates and at one individual missing a year of harvest.

```{r reorder}
Samples <- read.csv("c:/users/paul.conn/git/ckmr/bearded_adfg/Samples.csv")
head(Samples)

matches_POP <- read.csv("c:/users/paul.conn/git/ckmr/bearded_adfg/POP_matches.csv")
matches_MHS <- read.csv("c:/users/paul.conn/git/ckmr/bearded_adfg/MHS_matches.csv")
matches_PHS <- read.csv("c:/users/paul.conn/git/ckmr/bearded_adfg/PHS_matches.csv")
matches_POP$IDi = Samples$Our_sample[matches_POP$i]
matches_POP$IDj = Samples$Our_sample[matches_POP$j]
matches_MHS$IDi = Samples$Our_sample[matches_MHS$i]
matches_MHS$IDj = Samples$Our_sample[matches_MHS$j]
matches_PHS$IDi = Samples$Our_sample[matches_PHS$i]
matches_PHS$IDj = Samples$Our_sample[matches_PHS$j]


Which_remove = c(unique(grep("LowGene",Samples$Aging.comments)),which(is.na(Samples$YEAR)))
Samples_new = Samples[-Which_remove,]

#now reorder i,j based on those remaining
for(irec in 1:nrow(matches_POP)){
  matches_POP$i[irec] = which(Samples_new$Our_sample %in% matches_POP$IDi[irec])
  matches_POP$j[irec] = which(Samples_new$Our_sample %in% matches_POP$IDj[irec])
}
for(irec in 1:nrow(matches_MHS)){
  matches_MHS$i[irec] = which(Samples_new$Our_sample %in% matches_MHS$IDi[irec])
  matches_MHS$j[irec] = which(Samples_new$Our_sample %in% matches_MHS$IDj[irec])
}
for(irec in 1:nrow(matches_PHS)){
  matches_PHS$i[irec] = which(Samples_new$Our_sample %in% matches_PHS$IDi[irec])
  matches_PHS$j[irec] = which(Samples_new$Our_sample %in% matches_PHS$IDj[irec])
}
  
```

## Newborn pups vs yearlings

One issue is that there are many individuals harvested in the spring that are recorded as age '0' but when examining further notes it is possible to determine that they are either weeks old or 11 months old.  Since we anticipate employing a population model with a postbreeding census with an annual time step that 
occurs on May 1, we'll want to adjust the month of harvest for the 'weeks old' ones so that they are harvested in May (to be associated with that years cohort), and to adjust the 11 month old ones so that they're harvested in April (so that they are associated with the previous years cohort).

First, we'll treat 0 year old observations recorded by Lara Horstmann as if they are truly "weeks old" since weeks vs. 11 month old determinations are not available for these.  Based on the time samples were collected, this is clearly the most likely option. Using this approach, here are eight newborn pups that are harvested in April - for these we will shift the month of harvest to May.  There are also 3 individuals harvested in May that have an age of "11 months" - we'll adjust these to have a harvest in April so that they're associated with the previous years cohort.

```{r newborns}
  Pup_data = read.csv("c:/users/paul.conn/git/CKMR/bearded_adfg/Spring_pups_info.csv",header=TRUE)
  Pup_data$Adjusted.aging.comments[which(Pup_data$Adjusted.aging.comments=="weeks")]="Weeks" 
  Which_Lara = grep("Lara",Pup_data$Adjusted.aging.comments)
  Pup_data[Which_Lara,"Adjusted.aging.comments"]="Weeks"
  Newborns_april = which(Pup_data$MONTH=="APR" & Pup_data$Adjusted.aging.comments=="Weeks")
  Which_new = which(Samples_new$Our_sample %in% Pup_data$Our_sample[Newborns_april])
  Samples_new[Which_new,"MONTH"]="MAY"
  Yearlings_may = which(Pup_data$MONTH=="MAY" & Pup_data$Adjusted.aging.comments=="11 months")
  Which_yearling = which(Samples_new$Our_sample %in% Pup_data$Our_sample[Yearlings_may])
  Samples_new[Which_yearling,"MONTH"]="APR"

```

## Uncertain months

In most cases a month of harvest was specified, but in others it was not.  Let's specify
a month for those that are missing or are inexact (e.g., "Winter").  Since our annual time step is May 1, we'll index May by month 0, and April by month 11.  Given this time step, we'll also associate harvests made January - April with the previous year. 

```{r date}
  Month = Samples_new$MONTH
  Month[which(Month=="Spring")]="MAY"
  Month[which(Month=="Summer")]="JULY"
  Month[which(Month=="Fall")]="SEP"
  Month[which(Month=="Winter")]="JAN"
  Month = replace(Month,Month=="MAY","0")
  Month = replace(Month,Month=="JUN","1")
  Month = replace(Month,Month=="JUL","2")
  Month = replace(Month,Month=="AUG","3")
  Month = replace(Month,Month=="SEP","4")
  Month = replace(Month,Month=="OCT","5")
  Month = replace(Month,Month=="NOV","6")
  Month = replace(Month,Month=="DEC","7")
  Month = replace(Month,Month=="JAN","8")
  Month = replace(Month,Month=="FEB","9")
  Month = replace(Month,Month=="MAR","10")
  Month = replace(Month,Month=="APR","11")
  Month = as.numeric(Month)
  mean_month=mean(Month,na.rm=TRUE)
  Month[which(is.na(Month))]=mean_month
  Samples_new$MONTH = Month
  
  Which_winter=which(Samples_new$MONTH > 7)
  Samples_new$YEAR[Which_winter]=Samples_new$YEAR[Which_winter]-1
```
  

## Age distributions

For each harvested seal, we either need to know its age exactly (for models where age is assumed known), or to specify probabilities that an animal is a particular age (for models where age is assumed uncertain).  Let's assign definitive ages as well as probability distributions for both types of analyses.

### Certain ages

The 'AgeCombined' field in 'Samples' has mostly been filled in with the best age estimate possible, but there are a few exceptions, and a few ages that are clearly wrong.  First, one of the POPs indicates that an age 1 male, harvested in 2015, was an offspring of a mother who was killed in 2013.  Clearly, this male (seal EB15PH003) must have been at least a two year old.  We'll thus change it to be 2 instead of 1. Second, one of the maternal half-sibling pairs (siblings with the same mother) had two individuals with the same apparent birth years.  Since bearded seals only have one pup per year,
this is clearly an error of some kind: either an aging error, or the two are actually a paternal HSP that are "lucky" enough to share mitochondrial DNA.  Of the two, I suspect an aging error is the most probable; for purposes of this particular analysis, we'll change the age of the 6 year old to be 7.  

There are also a number of missing AgeCombined values or entries that had missing or *very* uncertain ages (e.g., tooth aging failed and there was only claws to age with which are problematic for older age classes). Although it seems possible to come up with probability distributions for these animals, deleting them is probably the best choice for an analysis relying on age being certain. There are also about 10 records where the ages were 0 or 1, with the note "age 1 if DOK spring" - these are seals for which month of kill is unknown.  However, since most seals are harvested in the spring or early summer, it is likely that these animals are one year olds (rather than a young-of-year harvested in the winter that might start to have an annuli that looks like a one year old). Lets change these to one-year-olds, get rid of the really uncertain records, make sure none are involved in kin-pair matches, and readjust kin-pair entries.

```{r certain_age}
  Samples_certain_age <- Samples_new
  Samples_certain_age$AgeCombined[which(Samples_new$Our_sample=='EB15PH003')]=2
  Samples_certain_age$AgeCombined[which(Samples_new$Our_sample=='EB13PH016')]=7
  
  Which_01 = which(Samples_certain_age$AgeCombined==0.5)
  Samples_certain_age$AgeCombined[Which_01]=1
  
  Which_remove = c(which(is.na(Samples_certain_age$AgeCombined)),
                   which((Samples_certain_age$AgeCombined %% 1)!=0))  #get rid of 0.5 age increments which were used to diagnose aging issues
  
  Remove_IDs = Samples_certain_age$Our_sample[Which_remove]
  Kin_pair_IDs = c(matches_POP$IDi,matches_POP$IDj,matches_MHS$IDi,matches_MHS$IDj,
                   matches_PHS$IDi,matches_PHS$IDj)
  which(Remove_IDs %in% Kin_pair_IDs)  #none involved in a kin pair
  Samples_certain_age = Samples_certain_age[-Which_remove,]
  
  matches_POP_ca = matches_POP
  matches_MHS_ca = matches_MHS
  matches_PHS_ca = matches_PHS
  
  #now reorder i,j based on those remaining
  for(irec in 1:nrow(matches_POP)){
    matches_POP_ca$i[irec] = which(Samples_certain_age$Our_sample %in% matches_POP$IDi[irec])
    matches_POP_ca$j[irec] = which(Samples_certain_age$Our_sample %in% matches_POP$IDj[irec])
  }
  for(irec in 1:nrow(matches_MHS)){
    matches_MHS_ca$i[irec] = which(Samples_certain_age$Our_sample %in% matches_MHS$IDi[irec])
    matches_MHS_ca$j[irec] = which(Samples_certain_age$Our_sample %in% matches_MHS$IDj[irec])
  }
  for(irec in 1:nrow(matches_PHS)){
    matches_PHS_ca$i[irec] = which(Samples_certain_age$Our_sample %in% matches_PHS$IDi[irec])
    matches_PHS_ca$j[irec] = which(Samples_certain_age$Our_sample %in% matches_PHS$IDj[irec])
  }
  
  #list object to export for further analysis
  CKMR_certain_age = list(Samples=Samples_certain_age,matches_POP=matches_POP_ca,
                          matches_MHS=matches_MHS_ca, matches_PHS=matches_PHS_ca)

```


### Uncertain ages analysis

Although a "certain age" analysis will be informative, it presumably won't do a very good job in
propagating uncertainty about ages into resulting abundance estimates.  For this reason, our next
step is to output probability distributions that represent the likely range of age estimates for each animal.  

First, seals with aged teeth were given a uniform distribution over the age range provided by Matson's aging laboratory; however, we admitted the possibility that animals could be up to one year younger or older than this range because our POP data indicated at least one case where the Matson age range was wrong by a year.  Specifically, we set the probability mass for aging one year younger or one year greater equal to 5% of the probability mass associated with the Matson age range.  

Second, there were a number of seals that did not have tooth readings but who were aged as pups using claws.  Through an auxiliary study that examined paired tooth and claw readings, pups that had claw age readings of zero we set $\pi_{i,0}=0.85$ and $\pi_{i,1}=0.15$ (that is, we allowed for a 15% probability that these animals were actually one year old).

Finally, we set a number of other seal age distributions on a case by case basis:

* Seal EB20PH002 had two teeth aged, which came back is 2 and 5-6.  For this seal, we used a uniform distribution on [2,6].  

* 13 seals with teeth aged as "0 or 1" were treated as seals that had an age range of 0 to 1 

* Two seals (EB17GAM027 and EB17PH035) had claw ages of 7+ but ambiguous tooth ages; these were assigned a uniform age distribution on [7,40] (40 was the maximum age allowed in CKMR modeling).

* Two seals (EB15PH036 and EB14PH001) had claw ages of 8 or 8+; since claw ages tend to underestimate true ages for older animals, these were assigned a uniform distriubiton on [8,40]

* Ten seals had missing ages ("UNK12GAM014","DIO-045-02","SH-G4-98","2012BS11","2011BS24","SH-G5-98","2014BS22","2011BS37","GM-J02-99","EB16SH006a").  These were assigned uniform distributions on [0,40].

Let's put these ideas into action.

```{r uncertain_age}

  Samples_uncertain = Samples_new

  n_indiv = nrow(Samples_new)
  Age_prior = matrix(0,n_indiv,41)

  Which_tooth = which(Samples_new$AgeType=="Tooth")
  n_tooth = length(Which_tooth)
  
  Which_missing = which(is.na(Samples_uncertain$Age1))
  Samples_uncertain[Which_missing,"Age1"]=Samples_uncertain[Which_missing,"Age2"]=
    Samples_uncertain[Which_missing,"AgeCombined"]
  
  for(i in 1:n_tooth){
    age1 = Samples_uncertain[Which_tooth[i],"Age1"]
    age2 = Samples_uncertain[Which_tooth[i],"Age2"]
    nyrs = age2-age1+1
    if(nyrs==1 & age1==0){ #pup so can't be a year younger
      Age_prior[Which_tooth[i],1]=0.95
      Age_prior[Which_tooth[i],2]=0.05
    }
    else{  #older than pup
      pi_i = 1/(nyrs+.1)
      Age_prior[Which_tooth[i],1+c(age1:age2)]=pi_i  #add one since ages go 0:40
      Age_prior[Which_tooth[i],age2+2]=pi_i*0.05
      Age_prior[Which_tooth[i],age1]=pi_i*0.05
    }
  }
  
  #age 0 claw ages
  Which_claw0 = which(Samples_new$AgeType=="Claw" & Samples_new$AgeCombined==0)
  Age_prior[Which_claw0,1]=0.85
  Age_prior[Which_claw0,2]=0.15
  
  #age 0.5 claw ages (equally likely 0 or 1)
  Which_claw = which(Samples_new$AgeType=="Claw" & Samples_new$AgeCombined==0.5)
  Age_prior[Which_claw,1]=0.5
  Age_prior[Which_claw,2]=0.5
  
  #age 7.5
  Which_7 = which(Samples_new$AgeCombined==7.5)
  Age_prior[Which_7,8:41]=1/34
  
  #age 8.5
  Which_8 = which(Samples_new$AgeCombined==8.5)
  Age_prior[Which_8,9:41]=1/33
  
  #unknown age
  Which_na = which(is.na(Samples_new$AgeCombined))
  Age_prior[Which_na,1:41]=1/41
  
  #check to see that all rows sum to one
  sum(Age_prior)==n_indiv  #should equal n_indiv 
```

Let's take a look at an example of an aging error "prior."  For example, here's a seal aged with a tooth with a range of 5-6 and a harvest date in July.  Here, most mass is on ages 5 and 6, but there is still a small possibility the seal is aged at 4 or 7.

```{r hist1}
plot(c(0:40),Age_prior[1,],xlab="Age",ylab="Probability mass")
```

Finally, we'll turn these data into a list for CKMR estimation:

```{r output}

  CKMR_uncertain_age = list(Samples=Samples_uncertain,matches_POP=matches_POP,
                          matches_MHS=matches_MHS, matches_PHS=matches_PHS,Age_prior=Age_prior)

  save(CKMR_certain_age,CKMR_uncertain_age,file="CKMR_sample_data.RData")

```


## A note on prior distributions based on age structure

So far, we have completely ignored the issue of prior distributions based on age structure.  For instance, if we knew nothing about an animals age, it would be much more likely for it to be a young animal than and old animal, simply because older animals must run the gauntlet of survival.  In this case, it would make sense to construct a prior distribution for age where $\pi_a$ (the probability an animal is age $a$) is proportional to $l(a)$, or the probability that animal survives to be age $a$.  However, we prefer not to add this element in to our prior distribution calculations at this stage because we anticipate that survivorship will be partly informed by half-sib data. That is, HSPs will inform the estimation of survival, so that it makes the most sense to do this adjustment within CKMR estimation routines.  In essence, in this writeup we have conducted an analysis with a uniform prior on age, whereas our desired prior depends on survival. We anticipate CKMR models will internally update our priors using

$$ [Age_i | y_i] \propto [y_i | Age_i] [Age_i] $$ 

where $y_i$ gives the aging data used in this writeup, $[y_i | Age_i]$ gives the probability distributions generated in this writeup, and $Age_i$ gives a survival-based prior distribution.



